---
title: "County-Level Ag and Observation Analysis"
output: html_notebook
---
<!--
Provenance: copied from `Raw Data/Bridging Gaps Dataset/Bridging Gaps.Rmd` (last modified 2025-09-17) into the
Citizen Science on the Farm replication package, September 2026.
Edits at packaging: hard-coded machine paths/credentials removed; logic unchanged.
Main 2023 notebook: ag_raster, obs_raster, exact_extract county metrics, coverage, socioeconomic GLM.
-->
# Import Libraries
```{r}
library(terra)
library(sf)
library(exactextractr)
library(arrow)
library(dplyr)
```

# Prepare base data
```{r}
# Counties (already transformed to EPSG:5070)
counties_sf <- st_read("../US_Counties/US_Counties_SocioEconomics.gpkg",
                       layer = "US_Counties_Data") %>%
  st_transform(5070)

# NLCD raster (already in 5070)
nlcd <- rast("../NLCD2023/Annual_NLCD_LndCov_2023_CU_C1V0.tif")

# Check
print(nlcd)
```
# Create ag_raster (1 = ag, 0 = non-ag)
```{r}
# NLCD codes 81 and 82 = ag
ag_raster <- classify(nlcd, cbind(c(81, 82), 1), others = 0)
writeRaster(ag_raster, "../NLCD2023/ag_raster.tif", overwrite = TRUE)
```

# Initialize obs_raster
```{r}
library(terra)

# Load previously created ag_raster
ag_raster <- rast("../NLCD2023/ag_raster.tif")

# Initialize empty obs_raster with same dimensions and CRS
obs_raster <- rast(ag_raster)
values(obs_raster) <- 0
```

# Splitt parquet
```{r}
library(arrow)
library(dplyr)

# Read the giant Parquet file
gbif_big <- read_parquet("../GBIF_US_2023/GBIFUS2023.parquet")

# Remove missing coordinates (optional but recommended BEFORE splitting)
gbif_big <- gbif_big %>%
  filter(!is.na(decimalLongitude) & !is.na(decimalLatitude))

# How many total rows?
total_rows <- nrow(gbif_big)

# Batch settings
batch_size <- 5e6
num_batches <- ceiling(total_rows / batch_size)

cat("Splitting into", num_batches, "files.\n")

# Create folder for split files if it doesn't exist
dir.create("../GBIF_US_2023/split/", showWarnings = FALSE, recursive = TRUE)

# Write out multiple small Parquet files
for (i in 1:num_batches) {
  start <- (i - 1) * batch_size + 1
  end <- min(i * batch_size, total_rows)

  batch <- gbif_big[start:end, ]

  filename <- paste0("../GBIF_US_2023/split/GBIF_batch_", i, ".parquet")
  write_parquet(batch, filename)

  cat("✅ Wrote", filename, "\n")
}

cat("✅✅ Splitting complete.\n")
```

# Create the obs_raster in batches
```{r}
# List all split Parquet files
parquet_files <- list.files("../GBIF_US_2023/split/", pattern = "\\.parquet$", full.names = TRUE)

cat("📂 Found", length(parquet_files), "batch files.\n")

for (i in seq_along(parquet_files)) {

  cat("🔎 Processing file", i, "of", length(parquet_files), ":", basename(parquet_files[i]), "\n")

  # Read batch
  batch <- read_parquet(parquet_files[i])

  # Skip empty batch
  if (nrow(batch) == 0) {
    cat("⚠ Empty batch. Skipping.\n")
    next
  }

  # Convert to terra vect directly (no sf!)
  gbif_vect <- vect(cbind(batch$decimalLongitude, batch$decimalLatitude),
                    type = "points", crs = "EPSG:4326")

  # Reproject
  gbif_vect <- project(gbif_vect, crs(ag_raster))

  # Rasterize this batch
  batch_raster <- rasterize(gbif_vect, ag_raster, fun = "count", background = 0)

  # Add to the cumulative obs_raster
  obs_raster <- obs_raster + batch_raster

  cat("✅ Finished batch", i, "\n")
}

# Save final obs_raster
writeRaster(obs_raster, "../NLCD2023/obs_raster.tif", overwrite = TRUE)
cat("🎉 All batches complete. obs_raster saved.\n")
```

# Prepare data for Google Earth Engine
```{r}
library(arrow)

# List all split Parquet files
parquet_files <- list.files("../GBIF_US_2023/split/", pattern = "\\.parquet$", full.names = TRUE)

# Loop through and save CSVs with just coords
for (i in seq_along(parquet_files)) {

  batch <- read_parquet(parquet_files[i])

  # Drop missing coordinates
  batch <- batch[!is.na(batch$decimalLongitude) & !is.na(batch$decimalLatitude), ]

  # Keep just coords
  coords <- batch[, c("decimalLongitude", "decimalLatitude")]

  # Write CSV
  filename <- paste0("../GBIF_US_2023/split/GBIF_batch_", i, ".csv")
  write.csv(coords, filename, row.names = FALSE)

  cat("✅ Wrote", filename, "\n")
}
```

# Merge csv files
```{r}
library(data.table)

# List all CSVs
csv_files <- list.files("../GBIF_US_2023/split/", pattern = "\\.csv$", full.names = TRUE)

# Read and rbind all CSVs quickly
all_points <- rbindlist(lapply(csv_files, fread))

# Confirm structure
head(all_points)

# Write the merged CSV
fwrite(all_points, "../GBIF_US_2023/GBIF_ALL_POINTS.csv")

cat("✅ Merged CSV saved as GBIF_ALL_POINTS.csv\n")
```

# Google Earth Engine Code
// Load your uploaded GBIF observation raster
var obs_raster = ee.Image("projects/nicholas-gunner/assets/obs_raster");

// Create binary presence/absence
var obs_binary = obs_raster.gt(0);

// Distance to nearest observation
var distance = obs_binary.fastDistanceTransform(30).sqrt().multiply(30);

// Create binary 1km proximity raster
var obs_1km = distance.lte(1000).selfMask();

// Load NLCD 2021 (same grid as your ag_raster / NLCD 2023)
var nlcd = ee.Image("USGS/NLCD_RELEASES/2021_REL/NLCD/2021").select('landcover');

var projection = nlcd.projection();

// --- Manually set CRS because .crs() fails ---
var crs = 'EPSG:5070';

// Extract transform (this still works fine)
var transform = projection.getInfo().transform;

// Get region
var region = nlcd.geometry();

// Export snapped to NLCD grid
Export.image.toDrive({
  image: obs_1km,
  description: 'obs_1km_raster',
  folder: 'EarthEngineExports',
  scale: 30,
  crs: crs,
  crsTransform: transform,
  region: region,
  maxPixels: 1e13
});


# Merge Google Earth Engine 1km Buffer Tiffs
```{r}
library(terra)

# List the exported TIFFs
tif_files <- list.files("../obs_1km/", pattern = "\\.tif$", full.names = TRUE)

# Read the rasters
r_list <- lapply(tif_files, rast)

# Merge them even if extents/alignments don't match
obs_1km_raster <- do.call(merge, r_list)

# Optional: save the unified raster
writeRaster(obs_1km_raster, "obs_1km_raster_merged.tif", overwrite = TRUE)
```

# ANALYSIS!!

```{r}
library(terra)
library(sf)
library(dplyr)
library(exactextractr)  # For efficient zonal stats

# Load rasters
ag_raster <- rast("ag_raster.tif")
obs_raster <- rast("obs_raster.tif")
obs_1km_raster <- rast("obs_1km_raster_merged.tif")
crs(obs_1km_raster) <- crs(ag_raster)

# Load counties
counties_sf <- st_read("counties_with_data.geojson")

# Ensure all data in same CRS
crs_target <- crs(ag_raster)
counties_sf <- st_transform(counties_sf, crs_target)

# Check alignment
stopifnot(all.equal(crs(obs_raster), crs(ag_raster)))
stopifnot(all.equal(crs(obs_1km_raster), crs(ag_raster)))
```
## Ag areas and percentages of observation on ag calculations
perc_ag_area
% of county area classified as ag

perc_obs_on_ag
% of all GBIF observations in the county made on ag land

```{r}
# 1️ Percentage of county that is agricultural
perc_ag_area <- exact_extract(ag_raster, counties_sf, 'mean', progress = TRUE)

# 2️ Percentage of GBIF observations made on agricultural lands

# To calculate:
# - numerator = exact sum of obs_raster * ag_raster
# - denominator = exact sum of obs_raster

ag_obs_sum <- exact_extract(obs_raster * ag_raster, counties_sf, 'sum', progress = TRUE)
total_obs_sum <- exact_extract(obs_raster, counties_sf, 'sum', progress = TRUE)

perc_obs_on_ag <- ag_obs_sum / total_obs_sum
```
```{r}
counties_sf$perc_ag_area <- perc_ag_area
counties_sf$perc_obs_on_ag <- perc_obs_on_ag
```

## Coverage Calculations
ag_observational_coverage
% of ag land within 1km of any observation
```{r}
# 3️ Agricultural observational coverage (% of ag land *not* within 1km of obs → so 1 - coverage)
# Ensure obs_1km_raster is aligned to ag_raster grid
obs_1km_raster <- resample(obs_1km_raster, ag_raster, method = "near")
# Mean of obs_1km_raster restricted to ag_raster
obs_1km_ag_mean <- exact_extract(obs_1km_raster * ag_raster, counties_sf, 'mean', progress = TRUE)

# Observational coverage = proportion of ag land within 1km of an observation
ag_observational_coverage <- obs_1km_ag_mean
```

```{r}
counties_sf$ag_observational_coverage <- ag_observational_coverage
```

## Write out final datasets

```{r}
st_write(counties_sf, "counties_with_final_metrics.geojson", delete_dsn = TRUE)
```

# Socioeconomic analysis

```{r}
str(counties_sf)
```

```{r}
# Convert some variables to numbers
counties_sf$POP_ESTIMATE_2023 <- as.numeric(gsub(",", "", counties_sf$POP_ESTIMATE_2023))
counties_sf$Median_Household_Income_2022 <- as.numeric(gsub(",", "", counties_sf$Median_Household_Income_2022))
```

```{r}
counties_sf$POP_ESTIMATE_2023 <- as.numeric(gsub(",", "", counties_sf$POP_ESTIMATE_2023))
counties_sf$Median_Household_Income_2022 <- as.numeric(gsub(",", "", counties_sf$Median_Household_Income_2022))

# Simple GLM
model <- glm(
  ag_observational_coverage ~ perc_ag_area + POP_ESTIMATE_2023 +
    PCTPOVALL_2023 + Percent.of.adults.with.a.bachelor.s.degree.or.higher..2019.23 +
    Unemployment_rate_2023 + Median_Household_Income_2022,
  data = counties_sf,
  family = quasibinomial()  # Because your response is a proportion between 0 and 1
)

summary(model)
```
```{r}
library(car)

vif_values <- vif(model)
print(vif_values)
```

```{r}

# Keep only complete cases used in the model
counties_modeldata <- counties_sf %>%
  filter(
    !is.na(perc_ag_area) &
    !is.na(POP_ESTIMATE_2023) &
    !is.na(PCTPOVALL_2023) &
    !is.na(Percent.of.adults.with.a.bachelor.s.degree.or.higher..2019.23) &
    !is.na(Unemployment_rate_2023) &
    !is.na(Median_Household_Income_2022) &
    !is.na(ag_observational_coverage)
  )

model <- glm(
  ag_observational_coverage ~ perc_ag_area + POP_ESTIMATE_2023 +
    PCTPOVALL_2023 + Percent.of.adults.with.a.bachelor.s.degree.or.higher..2019.23 +
    Unemployment_rate_2023 + Median_Household_Income_2022,
  data = counties_modeldata,
  family = quasibinomial()
)

counties_modeldata$predicted <- predict(model, type = "response")

plot(
  counties_modeldata$ag_observational_coverage,
  counties_modeldata$predicted,
  xlab = "Observed Coverage",
  ylab = "Predicted Coverage",
  main = "Observed vs Predicted Coverage",
  pch = 20,
  col = "blue"
)
abline(0, 1, col = "red")
```
```{r}
null_dev <- model$null.deviance
resid_dev <- model$deviance

pseudo_r2 <- 1 - (resid_dev / null_dev)
pseudo_r2
```

```{r}
cor(counties_modeldata$ag_observational_coverage, counties_modeldata$predicted)^2
```
```{r}
counties_modeldata$area_km2 <- st_area(counties_modeldata) / 1e6  # Convert from m² to km²
counties_modeldata$area_km2 <- as.numeric(counties_modeldata$area_km2)  # Ensure numeric

counties_modeldata$POP_DENSITY <- counties_modeldata$POP_ESTIMATE_2023 / counties_modeldata$area_km2

summary(counties_modeldata$POP_DENSITY)

counties_modeldata$LOG_POP_DENSITY <- log1p(counties_modeldata$POP_DENSITY)  # log(1 + x)

model2 <- glm(
  ag_observational_coverage ~ perc_ag_area + LOG_POP_DENSITY +
    PCTPOVALL_2023 + Percent.of.adults.with.a.bachelor.s.degree.or.higher..2019.23 +
    Unemployment_rate_2023,
  data = counties_modeldata,
  family = quasibinomial()
)

summary(model2)

counties_modeldata$predicted <- predict(model2, type = "response")

plot(
  counties_modeldata$ag_observational_coverage,
  counties_modeldata$predicted,
  xlab = "Observed Coverage",
  ylab = "Predicted Coverage",
  main = "Observed vs Predicted Coverage",
  pch = 20,
  col = "blue"
)
abline(0, 1, col = "red")

null_dev <- model2$null.deviance
resid_dev <- model2$deviance

pseudo_r2 <- 1 - (resid_dev / null_dev)
pseudo_r2
```
```{r}
observational_disparity <- counties_modeldata_ag5$perc_obs_on_ag / counties_modeldata_ag5$perc_ag_area
counties_modeldata_ag5$observational_disparity <- counties_modeldata_ag5$perc_obs_on_ag / counties_modeldata_ag5$perc_ag_area

counties_modeldata_ag5 <- counties_modeldata_ag5 %>%
  mutate(
    observational_disparity = ifelse(perc_ag_area > 0 & perc_obs_on_ag > 0,
                                     perc_obs_on_ag / perc_ag_area,
                                     NA)
  )

counties_modeldata_ag5 <- counties_modeldata %>%
  filter(perc_ag_area > 0.05)

nrow(counties_modeldata_ag5)

model_ag5 <- glm(
  observational_disparity ~ perc_ag_area + LOG_POP_DENSITY +
    PCTPOVALL_2023 +
    Unemployment_rate_2023,
  data = counties_modeldata_ag5,
  family = gaussian()
)

summary(model_ag5)

null_dev <- model_ag5$null.deviance
resid_dev <- model_ag5$deviance

pseudo_r2_ag5 <- 1 - (resid_dev / null_dev)
pseudo_r2_ag5

counties_modeldata_ag5$predicted <- NA
counties_modeldata_ag5$predicted[complete.cases(counties_modeldata_ag5[, c(
  "perc_ag_area", "LOG_POP_DENSITY", "PCTPOVALL_2023",
  "Unemployment_rate_2023", "observational_disparity"
)])] <- predict(model_ag5, type = "response")

plot(
  counties_modeldata_ag5$observational_disparity,
  counties_modeldata_ag5$predicted,
  xlab = "Observed Disparity",
  ylab = "Predicted Disparity",
  main = "Observed vs Predicted Observational Disparity (Ag >5%)",
  pch = 20,
  col = "darkgreen"
)
abline(0, 1, col = "red")

vif(model_ag5)

library(boot)
cv_error_ag5 <- cv.glm(counties_modeldata_ag5, model_ag5, K = 10)$delta[1]
cv_error_ag5
```

```{r}
# Install and load the sf package if you haven't already
install.packages("sf")
library(sf)

# Read the GeoJSON file (local or URL)
geojson_path <- "counties_with_final_metrics.geojson"
geo_data <- st_read(geojson_path)

# View the data
print(geo_data)

# Drop the geometry column and create a new data frame
non_spatial_df <- st_drop_geometry(geo_data)

# View the result
head(non_spatial_df)
```
```{r}
library(dplyr)

non_spatial_df <- non_spatial_df %>%
  rename(
    MEDIAN_HOUSEHOLD_INCOME_STATE_NORMALIZED = Med_HH_Income_Percent_of_State_Total_2022
  )
```
```{r}
# Remove one or more columns by name
non_spatial_df <- non_spatial_df %>%
  select(-ag_observational_coverage)
```
```{r}
sapply(non_spatial_df, typeof)
```

```{r}
non_spatial_df <- non_spatial_df %>%
  mutate(
    POP_ESTIMATE = as.numeric(gsub(",", "", POP_ESTIMATE))
  )
```
```{r}
continental_states <- c(
  "AL","AZ","AR","CA","CO","CT","DE","FL","GA","ID","IL","IN","IA",
  "KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV",
  "NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD",
  "TN","TX","UT","VT","VA","WA","WV","WI","WY"
)

library(dplyr)

non_spatial_df_contus <- non_spatial_df %>%
  filter(STATE %in% continental_states)
```

```{r}
non_spatial_df <- non_spatial_df %>%
  mutate(
    OBS_DISPARITY = ifelse(!is.na(PERC_AG_AREA) & PERC_AG_AREA > 0,
                           PERC_OBS_ON_AG / PERC_AG_AREA,
                           NA)
  )
write.csv(non_spatial_df, "us_counties_ag_observations.csv", row.names = FALSE)
```

```{r}
library(dplyr)

non_spatial_df_nona <- non_spatial_df %>%
  filter(!is.na(OBS_DISPARITY))

write.csv(non_spatial_df_nona, "us_counties_ag_observations.csv", row.names = FALSE)
```

# Create County Dataset with Area and Observation Count

## Load Required Libraries and Data

```{r}
# Load required libraries
library(terra)
library(sf)
library(dplyr)
library(exactextractr)
library(arrow)

# Load counties data
cat("Loading counties data...\n")
counties_sf <- st_read("../US_Counties/US_Counties_SocioEconomics.gpkg",
                       layer = "US_Counties_Data") %>%
  st_transform(5070)

# Load observation raster
cat("Loading observation raster...\n")
obs_raster <- rast("../NLCD2023/obs_raster.tif")

# Ensure counties are in the same CRS as the raster
counties_sf <- st_transform(counties_sf, crs(obs_raster))

cat("Data loaded successfully.\n")
```

## Calculate County Areas and Observation Counts

```{r}
# Calculate county areas
cat("Calculating county areas...\n")
counties_sf$area_m2 <- st_area(counties_sf)
counties_sf$area_km2 <- as.numeric(counties_sf$area_m2) / 1e6  # Convert to km²
counties_sf$area_ha <- as.numeric(counties_sf$area_m2) / 1e4   # Convert to hectares

# Calculate total observation count per county
cat("Calculating observation counts per county...\n")
obs_count <- exact_extract(obs_raster, counties_sf, 'sum', progress = TRUE)
counties_sf$total_obs_count <- obs_count

# Calculate observation density (observations per km²)
counties_sf$obs_density_km2 <- counties_sf$total_obs_count / counties_sf$area_km2

# Display summary statistics
cat("\nSummary Statistics:\n")
cat("Number of counties:", nrow(counties_sf), "\n")
cat("Total area (km²):", sum(counties_sf$area_km2, na.rm = TRUE), "\n")
cat("Total observations:", sum(counties_sf$total_obs_count, na.rm = TRUE), "\n")
cat("Mean county area (km²):", mean(counties_sf$area_km2, na.rm = TRUE), "\n")
cat("Mean observations per county:", mean(counties_sf$total_obs_count, na.rm = TRUE), "\n")
cat("Mean observation density (obs/km²):", mean(counties_sf$obs_density_km2, na.rm = TRUE), "\n")
```

## Create and Preview Dataset

```{r}
# Create a simplified dataset with area and observation metrics
county_area_obs_data <- counties_sf %>%
  select(
    FIPS,
    STATE,
    COUNTYNAME,
    area_km2,
    area_ha,
    total_obs_count,
    obs_density_km2,
    geometry
  )

# Display first few rows
cat("\nFirst 10 counties with area and observation data:\n")
print(head(st_drop_geometry(county_area_obs_data), 10))
```

## Save Dataset in Multiple Formats

```{r}
# Save the new county dataset with area and observation count
# Save as CSV (non-spatial)
county_area_obs_csv <- st_drop_geometry(county_area_obs_data)
write.csv(county_area_obs_csv, "county_area_obs_data.csv", row.names = FALSE)
cat("✅ Saved county_area_obs_data.csv\n")

# Save as GeoJSON (spatial)
st_write(county_area_obs_data, "county_area_obs_data.geojson", delete_dsn = TRUE)
cat("✅ Saved county_area_obs_data.geojson\n")

# Save as Parquet for efficient storage
write_parquet(county_area_obs_csv, "county_area_obs_data.parquet")
cat("✅ Saved county_area_obs_data.parquet\n")

cat("\n🎉 County dataset with area and observation count created successfully!\n")
cat("Files created:\n")
cat("- county_area_obs_data.csv (CSV format)\n")
cat("- county_area_obs_data.geojson (GeoJSON format)\n")
cat("- county_area_obs_data.parquet (Parquet format)\n")
```

```{r}
# Optimized script to calculate land cover percentages by county from NLCD 2023 data
# Author: AI Assistant
# Date: Generated for land cover analysis
# Approach: Memory-efficient processing with multi-threading

# Load required libraries
library(terra)
library(sf)
library(dplyr)
library(parallel)
library(doParallel)
library(foreach)

# Set working directory
# Load the NLCD 2023 raster
cat("Loading NLCD 2023 raster...\n")
nlcd <- rast("../NLCD2023/Annual_NLCD_LndCov_2023_CU_C1V0.tif")

# Load counties data
cat("Loading counties data...\n")
counties_sf <- st_read("../US_Counties/US_Counties_SocioEconomics.gpkg",
                       layer = "US_Counties_Data") %>%
  st_transform(crs(nlcd))

cat("Counties loaded:", nrow(counties_sf), "\n")

# Memory-efficient function to analyze land cover for a single county
analyze_county_landcover_optimized <- function(county_index, counties_sf, nlcd_path) {
  # Load the county and raster fresh in each worker
  county <- counties_sf[county_index, ]
  fips <- county$FIPS
  
  tryCatch({
    # Load raster fresh to avoid memory issues
    nlcd_raster <- rast(nlcd_path)
    
    # Crop the NLCD raster to the county extent
    county_nlcd <- crop(nlcd_raster, county)
    
    # Mask to the county boundary
    county_nlcd <- mask(county_nlcd, county)
    
    # Get all values in the county
    values <- values(county_nlcd)
    values <- values[!is.na(values)]
    
    # Clean up memory
    rm(county_nlcd)
    gc()
    
    if (length(values) == 0) {
      return(data.frame(
        FIPS = fips,
        PERCENT_DEVELOPED = NA,
        PERCENT_FOREST = NA,
        PERCENT_WATER = NA,
        TOTAL_PIXELS = 0
      ))
    }
    
    # Calculate percentages
    total_pixels <- length(values)
    
    # Developed land (codes 21-24)
    developed_pixels <- sum(values %in% c(21, 22, 23, 24))
    percent_developed <- developed_pixels / total_pixels
    
    # Forest land (codes 41-43)
    forest_pixels <- sum(values %in% c(41, 42, 43))
    percent_forest <- forest_pixels / total_pixels
    
    # Water (code 11)
    water_pixels <- sum(values == 11)
    percent_water <- water_pixels / total_pixels
    
    return(data.frame(
      FIPS = fips,
      PERCENT_DEVELOPED = percent_developed,
      PERCENT_FOREST = percent_forest,
      PERCENT_WATER = percent_water,
      TOTAL_PIXELS = total_pixels
    ))
    
  }, error = function(e) {
    cat("Error processing county", fips, ":", e$message, "\n")
    return(data.frame(
      FIPS = fips,
      PERCENT_DEVELOPED = NA,
      PERCENT_FOREST = NA,
      PERCENT_WATER = NA,
      TOTAL_PIXELS = 0
    ))
  })
}

# Set up parallel processing
cat("Setting up parallel processing...\n")
n_cores <- min(detectCores() - 1, 6)  # Conservative core usage
cat("Using", n_cores, "cores for parallel processing\n")

# Register parallel backend
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Export necessary objects to workers
clusterExport(cl, c("counties_sf"), envir = environment())

# Get number of counties
n_counties <- nrow(counties_sf)
cat("Processing", n_counties, "counties in parallel...\n")

# Process counties in parallel with progress tracking
cat("Starting parallel processing...\n")
start_time <- Sys.time()

# Process in smaller chunks for better memory management
chunk_size <- max(1, floor(n_counties / (n_cores * 2)))
chunks <- split(1:n_counties, ceiling(seq_along(1:n_counties) / chunk_size))

cat("Processing", length(chunks), "chunks of approximately", chunk_size, "counties each\n")

all_results <- list()
for (chunk_idx in 1:length(chunks)) {
  chunk_indices <- chunks[[chunk_idx]]
  cat("Processing chunk", chunk_idx, "of", length(chunks), 
      "(", length(chunk_indices), "counties)...\n")
  
  chunk_results <- foreach(i = chunk_indices, 
                         .combine = rbind,
                         .packages = c("terra", "sf"),
                         .errorhandling = "pass") %dopar% {
    analyze_county_landcover_optimized(i, counties_sf, "../NLCD2023/Annual_NLCD_LndCov_2023_CU_C1V0.tif")
  }
  
  all_results[[chunk_idx]] <- chunk_results
  
  # Force garbage collection between chunks
  gc()
}

# Combine all results
results <- do.call(rbind, all_results)

# Stop parallel cluster
stopCluster(cl)

end_time <- Sys.time()
processing_time <- end_time - start_time
cat("Parallel processing completed in", round(as.numeric(processing_time, units = "mins"), 2), "minutes\n")

# Check results
cat("Results summary:\n")
cat("Successfully processed:", sum(!is.na(results$PERCENT_DEVELOPED)), "counties\n")
cat("Failed to process:", sum(is.na(results$PERCENT_DEVELOPED)), "counties\n")

# Add the new columns to the counties data
counties_sf$PERCENT_DEVELOPED <- results$PERCENT_DEVELOPED
counties_sf$PERCENT_FOREST <- results$PERCENT_FOREST  
counties_sf$PERCENT_WATER <- results$PERCENT_WATER

# Display summary statistics
cat("\nLand Cover Summary Statistics:\n")
cat("Mean % developed:", round(mean(results$PERCENT_DEVELOPED, na.rm = TRUE) * 100, 2), "%\n")
cat("Mean % forest:", round(mean(results$PERCENT_FOREST, na.rm = TRUE) * 100, 2), "%\n")
cat("Mean % water:", round(mean(results$PERCENT_WATER, na.rm = TRUE) * 100, 2), "%\n")

# Show some examples
cat("\nFirst 10 counties with land cover percentages:\n")
example_data <- counties_sf %>%
  select(FIPS, STATE, COUNTYNAME, PERCENT_DEVELOPED, PERCENT_FOREST, PERCENT_WATER) %>%
  st_drop_geometry() %>%
  head(10)

print(example_data)

# Save the updated counties data
cat("\nSaving updated counties data...\n")
st_write(counties_sf, "counties_with_landcover_optimized.geojson", delete_dsn = TRUE)

# Also create a CSV version for easy merging
counties_csv <- counties_sf %>%
  select(FIPS, STATE, COUNTYNAME, PERCENT_DEVELOPED, PERCENT_FOREST, PERCENT_WATER) %>%
  st_drop_geometry()

write.csv(counties_csv, "counties_landcover_percentages_optimized.csv", row.names = FALSE)

cat("✅ Analysis complete!\n")
cat("Files created:\n")
cat("- counties_with_landcover_optimized.geojson (spatial data with land cover percentages)\n")
cat("- counties_landcover_percentages_optimized.csv (CSV for merging)\n")

# Display summary by state
cat("\nLand cover summary by state (top 10 by mean % developed):\n")
state_summary <- counties_sf %>%
  st_drop_geometry() %>%
  group_by(STATE) %>%
  summarise(
    mean_developed = mean(PERCENT_DEVELOPED, na.rm = TRUE) * 100,
    mean_forest = mean(PERCENT_FOREST, na.rm = TRUE) * 100,
    mean_water = mean(PERCENT_WATER, na.rm = TRUE) * 100,
    n_counties = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(mean_developed)) %>%
  head(10)

print(state_summary)


```

